GPU-Beschleunigung der Material Point Method

### 2019-01-29 mscthesis.tex
gao et al 2018

These descriptions are different but they will yield the same measurements when related correctly. I.e. both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different. In the reference confguration lookup is done at the initial position of the particle. In the current configuration lookup happens at the particle's world position. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

Parallel reduction schemes using warp shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}

@@

\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
The P2G-pull- and P2G-sync-transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region. The active particle region is the region that needs to get reduced. Therefore, it is sufficient for the P2G-sync-transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}. The P2G-pull-transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}. Both are achievable with a single level of parallel reduction (in contrast to a pyramid). This is the \textit{per block bin count} from now on that uses as the functional input for the parallel reduction:
\begin{enumerate}
  \item \texttt{unary\_op(count) = count} .
  \item \texttt{binary\_op(lc,rc) = max(lc,rc)}.
\end{enumerate}

@@

The \textbf{MPM-Transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
\begin{equation}
  m_{p}^{tot}= \sum_p m_p =  \sum_i m_i = m_{g}^{tot}.
\end{equation}
Following this procedure, the same should be true for the total linear momentum $\boldsymbol{p}^{tot}$:
\begin{equation}
  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}
\end{equation}
and also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
\begin{equation}
  \boldsymbol{L}_{p}^{tot}= \sum_p \boldsymbol{x}_p \times m_p \boldsymbol{v}_p =  \sum_i \boldsymbol{x}_i \times m_i \boldsymbol{v}_i = \boldsymbol{L}_{g}^{tot}.
\end{equation}
The test case spreads one million particles uniformly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}. A roundtrip consists of a P2G-transfer and a G2P-transfer for relevant variables. In this process the (random) velocities will get filtered. In PIC-Transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$. APIC-Transfers can conserve the angular momentum better as they define a local velocity field around the particle. Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.

The total errors vary little between the methods presented in chapter \ref{sec:transfers}. They are caused by floating-point inaccuracy. The order of execution on the GPU can vary. Thus the floating-point error will vary from execution to execution.
\begin{table}[t]
\centering
  \begin{tabular}{|l|r|r|}\hline
    Quantity& Abs. error  & Rel. error \\ \hline
    \multicolumn{3}{|c|}{PIC-Transfer}\\\hline
    $m^{tot}_i$ & 6.438     & 6.437e-06     \\\hline
    $\boldsymbol{p}^{tot}_i$  & (0.882,0.977,0.018)&(1.417,1.722,0.003)e-04 \\\hline
    $\boldsymbol{L}^{tot}_i$  & \textbf{(27.81,43.55,24.35)}&(0.024,0.036,0.273)\\\hline
    \multicolumn{3}{|c|}{APIC-Transfer}\\\hline
    $m^{tot}_i$ & 7.188     & 7.188e-06     \\\hline
    $\boldsymbol{p}^{tot}_i$  & (1.230,0.990,0.855)&(1.975,1.745,1.371)e-04 \\\hline
    $\boldsymbol{L}^{tot}_i$  & \textbf{(0.081,0.184,5.6e-04)}&(0.7043,1.510,0.063)e-04\\\hline
   \end{tabular}
   \caption{Error of 1,000 PIC/APIC-roundtrips of 1M uniformly placed particles with random velocity in a $128\times128\times128$ grid. Used this instance is a P2G-sync-transfer and G2P-transfer with batching = 4 and block size = (4,4,4). Particles are fixed.}
  \label{tab:roundtrip}
\end{table}
\clearpage
\subsection{Performance}\label{sec:perf}
The given metrics in \ref{sec:metrics} give valuable insight into how well the GPU can utilize a given shader or program. The GPU was a black box for a long time in this regard. The metrics provided by NVIDIA are a first step to open up the GPUs inner workings. Theory about optimization for the architecture such as \cite{NVIDIA:BEST:PRACTICE} can be more readily applied. However for abstraction purposes theory as well as descriptions about the architecture might be deliberately vague. These metrics give the programmer more power of identifying the right optimizations for a given code.

@@

Immediately apparent becomes that the SoA-Layout is 2.26 times faster. However, both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout. Though the big difference is in the memory access, the wrong access may be conversely spotted in the non-memory subunits. Since an AoS-Layout loads twice as many \texttt{vec4}, every non-memory related metric shrinks by half. An AoS-Layout however issues the same instructions. The SM should do equal work and be active for the same amount of cycles which is not the case.
\begin{table}[htpb]
  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
    Layout &  $\Delta t_c (\mu s)$ & Speedup       &VRAM             & L2              & L2-Hit & Read Active \\\hline
    AoS         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
    SoA         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
    \hline
  \end{tabular}
\caption{Metrics between different buffer layouts. Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length}\texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
\label{tab:map_2}
\end{table}

@@

The \textbf{P2G-sync-transfer} also may use batching and save time. The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}. Batching in this case reduces the amount of thread group barriers needed for synchronization. The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process. The external synchronization is however much less invasive than atomic writes. The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.
\begin{table}[htpb]
  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occup.\\\hline
    P2G-atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
    indirect		   &    6,276             & 1.10x   &90.8\%&\textbf{S. Scoreb.(31.7\%)} & 43.4\\\hline
    batching=4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& \textbf{34.9}\\\hline
    both                   &    3,148             & \textbf{2.20x}   &\textbf{65.0}\%&No Instr.(23.2\%) & 35.4\\\hline
    \hline
    P2G-sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & \textbf{35.4}\\\hline
 \end{tabular}
 \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. They form a rotated (unsorted) cube with four particles per cell. Block size is (8,4,4). Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:p2g_optim}
\end{table}

@@

However, before this process can be concluded merge and split techniques need to be taken into consideration. They are designed to augment the numerics of the material point method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp. As such they will stray less from the optimistic example. The extent to which they will stray from it dependends however on the dynamics of the problem.
\clearpage
\begin{appendices}
\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
  \begin{code}
\caption{$\boldsymbol{D}_p$ proof}
  \begin{minted}{python}
import numpy as np
from sympy import *

@@

\textbf{A Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid. This again forms a uniform grid. For simplicity assume two dimensions:

>>>>

These descriptions are different but they will yield the same measurements when related correctly. I.e. both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different. In the reference configuration lookup is done at the initial position of the particle. In the current configuration lookup happens at the particle's world position. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

Parallel reduction schemes using warp shuffle-operations may use sequential or interleaved addressing. Their stride is limited by the memory restrictions imposed due to a warp, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}

@@

\subsubsection{Active blocks \& maximum block count:}\label{sec:active_blocks}
The P2G-pull- and P2G-sync-transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region. The active particle region is the region that needs to get reduced. Therefore, it is sufficient for the P2G-sync-transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}. The P2G-pull-transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}. Both are achievable with a single level of parallel reduction (in contrast to a pyramid). This is the \textit{per block bin count} from now on that uses as the functional input for the parallel reduction:
\begin{enumerate}
  \item \texttt{unary\_op(count) = count} .
  \item \texttt{binary\_op(lc,rc) = max(lc,rc)}.
\end{enumerate}

@@

The \textbf{MPM-transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
\begin{equation}
  m_{p}^{tot}= \sum_p m_p =  \sum_i m_i = m_{g}^{tot}.
\end{equation}
Following this procedure, the same should be true for the total linear momentum $\boldsymbol{p}^{tot}$:
\begin{equation}
  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}
\end{equation}
and also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
\begin{equation}
  \boldsymbol{L}_{p}^{tot}= \sum_p \boldsymbol{x}_p \times m_p \boldsymbol{v}_p =  \sum_i \boldsymbol{x}_i \times m_i \boldsymbol{v}_i = \boldsymbol{L}_{g}^{tot}.
\end{equation}
The test case spreads one million particles uniformly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}. A roundtrip consists of a P2G-transfer and a G2P-transfer for relevant variables. In this process the (random) velocities will get filtered. In PIC-transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$. APIC-transfers can conserve the angular momentum better as they define a local velocity field around the particle. Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.

The total errors vary little between the methods presented in chapter \ref{sec:transfers}. They are caused by floating-point inaccuracy. The order of execution on the GPU can vary. Thus the floating-point error will vary from execution to execution.
\begin{table}[t]
\centering
  \begin{tabular}{|l|r|r|}\hline
    Quantity& Abs. error  & Rel. error \\ \hline
    \multicolumn{3}{|c|}{PIC-Transfer}\\\hline
    $m^{tot}_i$ & 6.438     & 6.437e-06     \\\hline
    $\boldsymbol{p}^{tot}_i$  & (0.882,0.977,0.018)&(1.417,1.722,0.003)e-04 \\\hline
    $\boldsymbol{L}^{tot}_i$  & \textbf{(27.81,43.55,24.35)}&(0.024,0.036,0.273)\\\hline
    \multicolumn{3}{|c|}{APIC-Transfer}\\\hline
    $m^{tot}_i$ & 7.188     & 7.188e-06     \\\hline
    $\boldsymbol{p}^{tot}_i$  & (1.230,0.990,0.855)&(1.975,1.745,1.371)e-04 \\\hline
    $\boldsymbol{L}^{tot}_i$  & \textbf{(0.081,0.184,5.6e-04)}&(0.7043,1.510,0.063)e-04\\\hline
   \end{tabular}
   \caption{Error of 1,000 PIC/APIC-roundtrips of 1M uniformly placed particles with random velocity in a $128\times128\times128$ grid. Used this instance is a P2G-sync-transfer and G2P-transfer with batching = 4 and block size = (4,4,4). Particles are fixed.}
  \label{tab:roundtrip}
\end{table}
\subsection{Performance}\label{sec:perf}
The given metrics in \ref{sec:metrics} give valuable insight into how well the GPU can utilize a given shader or program. The GPU was a black box for a long time in this regard. The metrics provided by NVIDIA are a first step to open up the GPUs inner workings. Theory about optimization for the architecture such as \cite{NVIDIA:BEST:PRACTICE} can be more readily applied. However for abstraction purposes theory as well as descriptions about the architecture might be deliberately vague. These metrics give the programmer more power of identifying the right optimizations for a given code.

@@

Immediately apparent becomes that the SoA-Layout is 2.26 times faster. However, both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout. Though the big difference is in the memory access, the wrong access may be conversely spotted in the non-memory subunits. Since an AoS-Layout loads twice as many \texttt{vec4}, every non-memory related metric shrinks by half. An AoS-Layout however issues the same instructions. The SM should do equal work and be active for the same amount of cycles which is not the case.
\begin{table}[htpb]
  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
    Layout &  $\Delta t_c (\mu s)$ & Speedup       &VRAM             & L2              & L2-Hit & Read Active \\\hline
    AoS         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
    SoA         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
    \hline
  \end{tabular}
\caption{Metrics between different buffer layouts. Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length} \texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}\label{tab:map_2}
\end{table}

@@

\begin{table}[htpb]
  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occup.\\\hline
    P2G-atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
    indirect		   &    6,276             & 1.10x   &90.8\%&\textbf{S. Scoreb.(31.7\%)} & 43.4\\\hline
    batching=4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& \textbf{34.9}\\\hline
    both                   &    3,148             & \textbf{2.20x}   &\textbf{65.0}\%&No Instr.(23.2\%) & 35.4\\\hline
    \hline
    P2G-sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & \textbf{35.4}\\\hline
 \end{tabular}
 \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. They form a rotated (unsorted) cube with four particles per cell. Block size is (8,4,4). Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:p2g_optim}
\end{table}

The \textbf{P2G-sync-transfer} also may use batching and save time. The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}. Batching in this case reduces the amount of thread group barriers needed for synchronization. The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process. The external synchronization is however much less invasive than atomic writes. The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.

@@

However, before this process can be concluded merge and split techniques need to be taken into consideration. They are designed to augment the numerics of the material point method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp. As such they will stray less from the optimistic example. The extent to which they will stray from it depends however on the dynamics of the problem.

At the end of this thesis awareness was raised to the following publication \cite{Gao:2018} which was simultaneously being worked on. Their decision making is very much alike:

Counting/Histogram sort is also their sorting algorithm of choice in the fashion of \cite{NVIDIA:NNSEARCH}. Instead of reordering any variable, they choose per variable if an index or deep sort is required. Their binning also floors in the same fashion.

The domain decomposition consists of $4\times4\times4$ blocks, also one of the best options employed here. Instead of filtering the active blocks in the pipeline, they motivate this process earlier by employing a sparse grid structure with fast neighbor queries altogether.

Implemented is a quadratic interpolation instead of a cubic one. The domain shift thus becomes $\frac{h}{2}$ instead of $h$. Accounting for the larger cubic weighting support, uniform distribution works as fast as the one shown here.

Their main difference in computation is the use of the aforementioned NVIDIA exclusive shuffle operations in the fashion of \cite{NVIDIA:SHUFFLE}. These operations largely solve two of the main problems the transfers have.

Instead of needing to rely on shared memory for inter-thread communication, the shuffle operation may be used. Each particle is assigned to a single thread. If for instance the particle size within the $4\times4\times4$ block exceeds the thread group size, another virtual block may be assigned to the same geometric block.

Each particle in a particle-to-grid-transfer then scatters their contribution to the neighboring grid nodes in a particle push fashion. The transfers are then solved due to segmented parallel reduction where the segments are the variable particle bin sizes. In shared memory this would cause numerous bank conflicts in both the scattering as well as the segmented parallel reduction as elucidated with the block scan of chapter \ref{sec:red_scan}. The scatter write would also need to be atomic.

This implementation tried to avoid these steps and schedule over grid nodes. This however is the cause of thread divergence which slows down computation for varying bin sizes. Therefore, \cite{Gao:2018} will achieve more throughput and better performance in these cases in exchange for a (at the moment) hardware vendor exclusive operation.

\clearpage
\begin{appendices}
\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
  \begin{code}
\caption{$\boldsymbol{D}_p$ proof}
  \begin{minted}{python}
import numpy as np
from sympy import *

@@

\textbf{Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid. A local scan of an n-dimensional tile/block in that grid can be computed based on this global scan. This again forms a uniform grid. For simplicity assume two dimensions:
